Survival Compass

### 2024-04-21 paper/paper.qmd
updated results section and made all graphs consistent in terms of display.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-one
#| fig-cap: Distribution of survival time in patients

@@

# Histogram of survival time in days (days_to_death)
ggplot(analysis_data, aes(x = days_to_death)) +
  geom_histogram(binwidth = 50, fill = "forestgreen", color = "black") +
  labs(title = "Distribution of Survival Time in Lung Cancer Patients Post-Diagnosis",
       x = "Days to Death", y = "Frequency") +
  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
                                  by = 500)) +  # Set breaks every 500 days
  theme_minimal() +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

### pathologic Stage
Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging. As depicted in Graph 1 of @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis. The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB. At this stage, the cancer is still found early however the tumor has regional spread to nearby lymph nodes. Interestingly, at diagnosis, patients had a fairly even spread across the different stages.

### Presence of Distant Metastasis
The presence of distant metastasis, indicative of cancer spread beyond the site of origin to nearby lymph nodes, significantly influences treatment decisions and patient outcomes. Graph 2 of @fig-two illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients. The figure indicates that when patients are diagnosed, most are found in category 'M0' having no distant metastasis and thus cancer has not spread beyond the site of origin. It is important to note that category 'MX', where tumor could not be evaluated has the next largest proportion of patients. Additionally, category 'M1a' appears to show no data however this is because only 1 out of the 382 patients were classified this way.

### Lymph Node Involvement
Lymph node involvement is another key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes. Graph 3 of @fig-two presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis. Here, there is a clear trend that as lymph node involvement becomes more severe, the proportion of patients decline. This demonstrates that upon diagnosis, most patients fall under the category 'N0' where there is no regional lymph node metastasis hence cancer has not spread to the lymph nodes.

### Tumor Size
Tumor size is measured based on the diameter of the primary tumor in lung cancer patients. Graph 4 of @fig-two showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis. The highest proportion of patients were categorized as 'T2' where the tumor is between 3 to 5 centimeters in diameter has grown into the inner lining of the lung, possibly leads to swelling and or collapse of the lung. This category was overwhelmingly more prominent than all others by twofold. All other categories were relatively evenly distributed.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-two
#| fig-title: Percentage of Lung Cancer Patients by
#| fig-cap: Percentage of Lung Cancer Patients by Pathologic Stage, Presence of Distant Metastasis, Lymph Node Involvement, Tumor Size

@@

@@

# Create subplots for all four plots
plots <- list(
  ggplot(pathologic_stage_counts, aes(x = pathologic_stage, y = percentage)) +
    geom_bar(stat = "identity", fill = "lightpink") +
    labs(title = "Graph 1: Pathologic Stage",
         x = "",
         y = "Percentage") +
    theme_minimal() +
    theme(
      text = element_text(family = "patrick"), # Set font to Patrick Hand
      axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels by 45 degrees
    ) +
    scale_y_continuous(breaks = seq(0, max(pathologic_stage_counts$percentage), by = 5)),

   ggplot(presence_of_distant_metastasis_counts, aes(x = presence_of_distant_metastasis, y = percentage)) +
    geom_bar(stat = "identity", fill = "gold") +
    labs(title = "Graph 2: Presence of Distant Metastasis",
         x = "",
         y = "") +  # Remove y-axis label 
    theme_minimal() +
    theme(
      text = element_text(family = "patrick"), # Set font to Patrick Hand
      axis.title.y = element_blank(), # Remove y-axis title
    ),

   ggplot(lymph_node_involvement_counts, aes(x = lymph_node_involvement, y = percentage)) +
    geom_bar(stat = "identity", fill = "cyan") +
    labs(title = "Graph 3: Lymph Node Involvement",
         x = "",
         y = "Percentage") +  # Keep y-axis label for this plot
    theme_minimal() +
    theme(
      text = element_text(family = "patrick") # Set font to Patrick Hand
    ) +
    scale_y_continuous(breaks = seq(0, max(lymph_node_involvement_counts$percentage), by = 10)),

  ggplot(tumor_size_counts, aes(x = tumor_size, y = percentage)) +
    geom_bar(stat = "identity", fill = "purple") +
    labs(title = "Graph 4: Tumor Size",
         x = "",
         y = "") +  # Remove y-axis label 
    theme_minimal() +
    theme(
      text = element_text(family = "patrick"), # Set font to Patrick Hand
      axis.title.y = element_blank() # Remove y-axis title
    )
)

# Combine plots using patchwork and largen the dimensions
plots_combined <- wrap_plots(plots, ncol = 2, height = 12, width = 14)

@@

@@

Our results are summarized in @.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# install.packages
# install.packages("reshape2")
# install.packages("vcd")

# Load required libraries
library(ggplot2)
library(arrow)
library(dplyr)
library(reshape2)
library(vcd)

@@

# Create a contingency table
contingency_table <- table(analysis_data$pathologic_stage, 
                           analysis_data$presence_of_distant_metastasis, 
                           analysis_data$lymph_node_involvement, 
                           analysis_data$tumor_size, 
                           analysis_data$treatment_type)

# Plot mosaic plot for pathologic_stage and treatment_type with y-axis labels on the right side
mosaic(~pathologic_stage + treatment_type, data = analysis_data, shade = TRUE, labeling_args = list(rot_labels = c(90, 0)), gp = gpar(mar = c(5, 5, 5, 5)), split_vertical = TRUE)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Load required libraries
library(ggplot2)
library(arrow)

@@

# Scatter plot survival time vs. pathologic stage - to compare survival probabilities
ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
  geom_point(size = 3, color = "lightpink") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Pathologic Stage Post-Diagnosis",
       x = "Pathologic Stage",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

# Line graph with trend lines 
ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
  geom_line(color = "blue", group = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Days to Death vs. Pathologic Stage for Lung Cancer Patients",
       x = "Pathologic Stage",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# not sure what this is 
# Plotting
ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
  geom_density(alpha = 0.5) +
  geom_density(color = "black", size = 1.5) +
  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
       x = "Days to Death",
       y = "Density") +
  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
  theme_minimal()

# not sure .2
ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
  geom_density(alpha = 0.5) +
  geom_density(color = "black", size = 1.5) +
  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
       x = "Days to Death",
       y = "Density") +
  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
  theme_minimal() +
  facet_wrap(~pathologic_stage, scales = "free")

# Boxplot 
ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survial Distribution Post-Diagnosis by Pathologic Stage",
       x = "Pathologic Stage",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

######################################

# Scatter plot survival time vs. Presence of Distant Metastasis
ggplot(analysis_data, aes(x = presence_of_distant_metastasis, y = days_to_death)) +
  geom_point(size = 3, color = "gold") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Presence of Distant Metastasis Post-Diagnosis",
       x = "Presence of Distant Metastasis",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

# Scatter plot survival time vs. Lymph node involvement
ggplot(analysis_data, aes(x = lymph_node_involvement, y = days_to_death)) +
  geom_point(size = 3, color = "cyan") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Lymph Node Involvement Post-Diagnosis",
       x = "Lymph Node Involvement",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

# Scatter plot survival time vs. Tumor Size
ggplot(analysis_data, aes(x = tumor_size, y = days_to_death)) +
  geom_point(size = 3, color = "purple") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Tumor Size Post-Diagnosis",
       x = "Tumor Size",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

#########################

# Scatter plot survival time vs. Treatment Type
ggplot(analysis_data, aes(x = treatment_type, y = days_to_death)) +
  geom_point(size = 3, color = "cyan") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Treatment Type Post-Diagnosis",
       x = "Treatment Type",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

# Grouped bar chart for survival time vs. Treatment Type
ggplot(analysis_data, aes(x = treatment_type, y = days_to_death, fill = treatment_type)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Average Survival Time by Treatment Type Post-Diagnosis",
       x = "Treatment Type",
       y = "Average Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Horizontal grouped bar chart for survival time vs. Treatment Type
ggplot(analysis_data, aes(y = treatment_type, x = days_to_death, fill = treatment_type)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge", orientation = "horizontal") +
  labs(title = "Average Survival Time by Treatment Type Post-Diagnosis",
       y = "Treatment Type",
       x = "Average Survival Time (days)") +
  theme_minimal() +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Grouped dot plot for survival time vs. Treatment Type
ggplot(analysis_data, aes(x = treatment_type, y = days_to_death, color = treatment_type)) +
  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.7) +
  labs(title = "Survival Time by Treatment Type Post-Diagnosis",
       x = "Treatment Type",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Grouped dot plot with flipped axes for survival time vs. Treatment Type
ggplot(analysis_data, aes(y = treatment_type, x = days_to_death, color = treatment_type)) +
  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.7) +
  labs(title = "Survival Time by Treatment Type Post-Diagnosis",
       y = "Treatment Type",
       x = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1)) +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Stacked bar chart for treatment method vs. Average Survival Time
ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
       x = "Average Survival Time (days)",
       y = "Proportion of Patients") +
  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
  theme_minimal() +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Stacked bar chart for treatment method vs. Average Survival Time
ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
       x = "Average Survival Time (days)",
       y = "Proportion of Patients") +
  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
  scale_x_continuous(expand = c(0, 0)) + # Adjust x-axis limits
  theme_minimal() +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Stacked bar chart for treatment method vs. Average Survival Time
ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
       x = "Average Survival Time (days)",
       y = "Proportion of Patients") +
  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
  scale_x_continuous(expand = c(0, 0), limits = c(0, max(analysis_data$days_to_death))) + # Adjust x-axis limits
  theme_minimal() +
  theme(text = element_text(family = "patrick")) 
  # Set font to Patrick Hand

# Boxplot 
ggplot(analysis_data, aes(x = treatment_type, y = days_to_death)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survial Distribution Post-Diagnosis by Treatment Type",
       x = "Treatment Type",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

>>>>

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-one
#| fig-cap: The histogram shows the distribution of survival time (days_to_death) in lung cancer patients post-diagnosis.

@@

# Histogram of survival time in days (days_to_death)
ggplot(analysis_data, aes(x = days_to_death)) +
  geom_histogram(binwidth = 50, fill = "forestgreen", color = "black") +
  labs(title = "Distribution of Survival Time Post-Diagnosis",
       x = "Days to Death", y = "Frequency") +
  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
                                  by = 500)) +  # Set breaks every 500 days
  theme_minimal() +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

### pathologic Stage
Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging. As depicted in @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis. The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB. At this stage, the cancer is still found early however the tumor has regional spread to nearby lymph nodes. Interestingly, at diagnosis, patients had a fairly even spread across the different stages.

### Presence of Distant Metastasis
The presence of distant metastasis, indicative of cancer spread beyond the site of origin to nearby lymph nodes, significantly influences treatment decisions and patient outcomes. @fig-two illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients. The figure indicates that when patients are diagnosed, most are found in category 'M0' having no distant metastasis and thus cancer has not spread beyond the site of origin. We decided to include the classification 'MX', where tumor could not be evaluated as it contained the next largest proportion of patients. If this data was omitted, there would be a large skew in the data. Additionally, category 'M1a' appears to show no data however this can be explained as only 1 out of the 382 patients were classified this way.

### Lymph Node Involvement
Lymph node involvement is another key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes. @fig-two presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis. Here, there is a clear trend that as lymph node involvement becomes more severe, the proportion of patients decline. This demonstrates that upon diagnosis, most patients fall under the category 'N0' where there is no regional lymph node metastasis hence cancer has not spread to the lymph nodes.

### Tumor Size
Tumor size is measured based on the diameter of the primary tumor in lung cancer patients. @fig-two showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis. The highest proportion of patients were categorized as 'T2' where the tumor is between 3 to 5 centimeters in diameter has grown into the inner lining of the lung, possibly leads to swelling and or collapse of the lung. This category was overwhelmingly more prominent than all others by twofold. All other categories were relatively evenly distributed.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-two
#| fig-cap: The histogram shows the percentage of lung cancer patients by pathologic stage, presence of distant metastasis, lymph node involvement, tumor size

@@

# Set overall title and caption
overall_title <- "Percentage of Lung Cancer Patients by:"

@@

# Create subplots for all four plots
plots <- list(
  ggplot(pathologic_stage_counts, aes(x = pathologic_stage, y = percentage)) +
    geom_bar(stat = "identity", fill = "lightpink") +
    labs(title = "Pathologic Stage",
         x = "",
         y = "Percentage") +
    theme_minimal() +
    theme(
      text = element_text(family = "patrick"), # Set font to Patrick Hand
      axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels by 45 degrees
    ) +
    scale_y_continuous(breaks = seq(0, max(pathologic_stage_counts$percentage), by = 5)),

   ggplot(presence_of_distant_metastasis_counts, aes(x = presence_of_distant_metastasis, y = percentage)) +
    geom_bar(stat = "identity", fill = "gold") +
    labs(title = "Presence of Distant Metastasis",
         x = "",
         y = "") +  # Remove y-axis label 
    theme_minimal() +
    theme(
      text = element_text(family = "patrick"), # Set font to Patrick Hand
      axis.title.y = element_blank(), # Remove y-axis title
    ),

   ggplot(lymph_node_involvement_counts, aes(x = lymph_node_involvement, y = percentage)) +
    geom_bar(stat = "identity", fill = "cyan") +
    labs(title = "Lymph Node Involvement",
         x = "",
         y = "Percentage") +  # Keep y-axis label for this plot
    theme_minimal() +
    theme(
      text = element_text(family = "patrick") # Set font to Patrick Hand
    ) +
    scale_y_continuous(breaks = seq(0, max(lymph_node_involvement_counts$percentage), by = 10)),

  ggplot(tumor_size_counts, aes(x = tumor_size, y = percentage)) +
    geom_bar(stat = "identity", fill = "purple") +
    labs(title = "Tumor Size",
         x = "",
         y = "") +  # Remove y-axis label 
    theme_minimal() +
    theme(
      text = element_text(family = "patrick"), # Set font to Patrick Hand
      axis.title.y = element_blank() # Remove y-axis title
    ) + guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
    # Adjust legend title position

)

# Combine plots using patchwork and set overall title and caption
plots_combined <- wrap_plots(plots, ncol = 2, heights = c(1.2, 1.2)) +
  plot_annotation(
    title = overall_title,
    theme = theme(
      text = element_text(family = "patrick", size = 14),  # Adjust caption size
      plot.caption = element_text(family = "patrick", size = 12, hjust = 0) 
      # Move caption to far right
    )
  )

@@

### Treatment Type
Treatment type analysis examines the distribution of lung cancer patients based on the type of treatment they received. @fig-three illustrates this distribution, with treatment type on the x-axis and the percentage of patients on the y-axis. Pharmaceutical Therapy and Radiation Therapy were the two primary treatment categories observed. The split between usage of these two treatment types was very minimal with pharmaceutical therapy at 50.3% and radiation therapy at 49.7%.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-three
#| fig-cap: The histogram shows the percentage of lung cancer patients by treamtent type.

# Create bar chart for treatment type with color coding
treatment_type_counts <- analysis_data %>%
  group_by(treatment_type) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Define colors for pharmaceutical therapy and radiation therapy
color_palette <- c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "green")

# Plotting with color coding and percentages
ggplot(treatment_type_counts, aes(x = treatment_type, y = percentage,
                                  fill = treatment_type,
                                  label = paste0(round(percentage, 1), "%"))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), family = "patrick",
            vjust = -0.5, size = 4) +  
  scale_fill_manual(values = color_palette) +
  labs(x = "Treatment Type", y = "Percentage of Patients",
       title = "Distribution of Lung Cancer Patients by Treatment Type") +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, max(treatment_type_counts$percentage), by = 5)) +  
  theme(text = element_text(family = "patrick", size = 14))

@@

Our results are summarized in @fig-four, @fig-five, @fig-six and @fig-seven. This analysis investigates lung cancer treatment outcomes given either radiation therapy or pharmaceutical therapy as well as pathologic stage, presence of distant metastasis, lymph node involvement and tumor size.

## Treatment Type by Pathologic Stage
This scatter plot @fig-four provides a view of how pathologic stage correlates with the time to death among lung cancer patients, differentiated by treatment type. Each dot represents an individual patient, color-coded to represent whether they received pharmaceutical therapy or radiation therapy. On the x-axis we have days to death after diagnosis and on the y-axis we have the pathologic stages. There are considerably fewer patients part of stages I, II, III, and IIB at diagnosis compared to the other ones. As the time from diagnosis to death increases along the x-axis, we observe a general trend downwards in pathologic stage, indicating disease progression over time. The trend lines fitted for each treatment type further highlight this progression, with radiation therapy showing a steeper decline compared to pharmaceutical therapy. 
This illustrates that upon diagnosis, patients with more advanced pathologic stages are associated with a shorter survival time. Additionally, this suggests that patients receiving pharmaceutical therapy tend to exhibit a slower advancement in pathologic stage over time, thus increasing their survival time more so than those undergoing radiation therapy.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-four
#| fig-cap: The scatterplot shows the relationship between pathologic stage, time to death, and treatment type in lung cancer patients post-diagnosis.

# Load required libraries
library(ggplot2)
library(arrow)
library(showtext)

@@

# Create a scatter plot with trend lines - pathologic stage vs. time vs. treatment type
# Create scatter plot with trend lines - pathologic stage vs. time vs. treatment type
plot1 <- ggplot(data = analysis_data, aes(x = days_to_death, y = pathologic_stage, color = treatment_type)) +
  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  # Adjusted point size and transparency
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type), alpha = 0.3) +  # Adjust trend line transparency
  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
                                "Radiation Therapy, NOS" = "green")) +
  labs(x = "Days to Death", y = "Pathologic Stage", color = "Treatment Type") +
  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death), by = 500)) + 
  # Set breaks on x-axis
  theme_minimal() +
  theme(text = element_text(family = "patrick", size = 14),  # Set font to Patrick Hand and increase the size
        legend.position = "bottom", legend.text = element_text(size = 12),
        plot.caption = element_text(hjust = 0, size = 14)) +  # Adjust legend position and text size
  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
  ggtitle("Pathologic Stage vs. Time vs. Treatment Type")  # Title for the plot

# Display the plot
plot1

@@

## Treatment Type by Presence of Distant Metastasis
In this scatter plot, @fig-five we examine the relationship between the presence of distant metastasis, time to death, and treatment type for lung cancer patients. Similarly to figure above, the dots on the plot represent individual patients, color-coded based on their treatment regimen. Days to death after diagnosis is on the x-axis and on the y-axis we have the degree of the presence of distant metastasis. The majority of patients are diagnosed at classification 'MO' and hence are caught before cancer is spread from the primary site of origin. As patients survival time increases, excluding  category 'MX', there is a decrease in the presence of distant metastasis, denoted by the downward trend in the scatter plot. Patients undergoing pharmaceutical therapy show a sharper decrease in distant metastasis compared to those receiving radiation therapy. 
The trend line for radiation therapy is nearly a plateau, insinuating that radiation therapy was more effective in prolonging survival. This suggests that pharmaceutical therapy may  correlate with a lower likelihood of survival given distant metastasis over time compared to radiation therapy. Additionally, patients associated with higher classification of metastasis appear to have shorter survival times.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-five
#| fig-cap: The scatterplot illustrates the relationship between distant metastasis, time to death, and treatment type in lung cancer patients post-diagnosis.

# Load required libraries
library(ggplot2)
library(arrow)
library(showtext)

@@

# Create a scatter plot with trend lines - presence of distant metastasis vs. time vs. treatment type
plot2 <- ggplot(data = analysis_data, aes(x = days_to_death, y = presence_of_distant_metastasis, color = treatment_type)) +
  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
              alpha = 0.3) +  
  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
                                "Radiation Therapy, NOS" = "green")) +
  labs(x = "Days to Death", y = "Presence of Distant Metastasis",
       color = "Treatment Type") +
  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
                                  by = 500)) +
  theme_minimal() +
  theme(text = element_text(family = "patrick", size = 14), 
        legend.position = "bottom", legend.text = element_text(size = 12),
        plot.caption = element_text(hjust = 0, size = 14)) +  
  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
  ggtitle("Presence of Distant Metastasis vs. Time vs. Treatment Type")

# Display the plot
plot2
```
## Treatment Type by Lymph Node Involvement 
This scatter plot  @fig-six explores the relationship between lymph node involvement, time to death, and treatment type among cancer patients. Like above, each dot symbolizes an individual patient, with colors indicating their treatment type; pharmaceutical therapy or radiation therapy. Days to death is on the x-axis and lymph node involvement classifications are on the y-axis. At diagnosis, most patients are classified under either NO, N1 or N2. As survival time increases, there is an observable decline in lymph node involvement, reflected by the downward trend in the scatter plot. The trend lines further illuminate this pattern, with patients undergoing pharmaceutical therapy exhibiting a slightly steeper decline in lymph node involvement over time compared to those receiving radiation therapy, however this is very minimal and insignificant difference between the two treatment types. 
This suggests that pharmaceutical therapy and radiation therapy are both equally associated with increasing survival time based on lymph node involvement. Moreover, patients associated with higher classification of involvement with lymph nodes appear to have shorter survival times.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-six
#| fig-caption: The scatterplot demonstrates the relationship between lymph node involvement, time to death, and treatment type in lung cancer patients post-diagnosis.

# Load required libraries
library(ggplot2)
library(arrow)
library(showtext)

# Create a scatter plot with trend lines - lymph node involvement vs. time vs. treatment type
plot3 <- ggplot(data = analysis_data, aes(x = days_to_death,
                                          y = lymph_node_involvement,
                                          color = treatment_type)) +
  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
              alpha = 0.3) +  
  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
                                "Radiation Therapy, NOS" = "green")) +
  labs(x = "Days to Death", y = "Lymph Node Involvement",
       color = "Treatment Type") +
  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
                                  by = 500)) +
  theme_minimal() +
  theme(text = element_text(family = "patrick", size = 14), 
        legend.position = "bottom", legend.text = element_text(size = 12),
        plot.caption = element_text(hjust = 0, size = 14)) +  
  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
  ggtitle("Lymph Node Involvement vs. Time vs. Treatment Type")

# Display the plot
plot3
```
## Treatment Type by Tumor Size
In this scatter plot, @fig-seven we investigate the relationship between tumor size, time to death, and treatment type for cancer patients. Each dot on the plot corresponds to an individual patient, categorized by their treatment regimen whether it is pharmaceutical therapy or radiation therapy. The x-axis represents the time from diagnosis until death and the y-axis is comprised of the tumor sizes. At diagnosis most patients are classified with a T2 tumour. As survival time increases, there is a noticeable reduction in tumor size, depicted by the downwards trajectory in the scatter plot. The trend lines further accentuate this trend, with patients undergoing radiation therapy demonstrating a more pronounced decrease in tumor size over time compared to those receiving pharmaceutical therapy. 
This suggests that patients associated with larger tumors tend to have shorter survival times. This suggests that pharmaceutical therapy might be associated with a slower tumor growth rate and hence be more efficient in treating lung cancer patients in terms of tumor size.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-seven
#| fig-caption: The scatterplot portrays the relationship between tumor size, time to death, and treatment type in lung cancer patients post-diagnosis.

# Load required libraries
library(ggplot2)
library(arrow)
library(showtext)

# Create a scatter plot with trend lines - tumor size vs. time vs. treatment type
plot4 <- ggplot(data = analysis_data, aes(x = days_to_death,
                                          y = tumor_size,
                                          color = treatment_type)) +
  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
              alpha = 0.3) +  
  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
                                "Radiation Therapy, NOS" = "green")) +
  labs(x = "Days to Death", y = "Tumor Size",
       color = "Treatment Type") +
  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
                                  by = 500)) +
  theme_minimal() +
  theme(text = element_text(family = "patrick", size = 14), 
        legend.position = "bottom", legend.text = element_text(size = 12),
        plot.caption = element_text(hjust = 0, size = 14)) +  
  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
  ggtitle("Tumor Size vs. Time vs. Treatment Type")

# Display the plot
plot4
